---
title: "Stats_Module2_quiz"
author: "OHfromH2O"
date: '2022-06-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### 1. Load the Montgomery and Pickrell eSet:
```{r echo=TRUE}
library(ballgown)
library(Biobase)
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```

### **Q1. What percentage of variation is explained by the 1st principal component in the data set if you**:  
1. Do **no transformations**?
2. **log2(data + 1) transform**?
3. **log2(data + 1) transform and subtract row means**?
```{r echo=TRUE, message=FALSE, warning=FALSE}
# No transformations
svd1 = svd(edata)
plot(svd1$d^2/sum(svd1$d^2),ylab="Percent Variance Explained_1",col=2)
a = svd1$d^2/sum(svd1$d^2)
a[1]
```
```{r echo=TRUE, message=FALSE, warning=FALSE}
# log2 transform
edata_log2 = log2(edata + 1)
svd2 = svd(edata_log2)
plot(svd2$d^2/sum(svd2$d^2),ylab="Percent Variance Explained_2",col=3)
b = svd2$d^2/sum(svd2$d^2)
b[1]
```
```{r echo=TRUE, message=FALSE, warning=FALSE}
# log2 transform and subtract row means
edata_centered = edata_log2 - rowMeans(edata_log2)
svd3 = svd(edata_centered)
plot(svd3$d^2/sum(svd3$d^2),ylab="Percent Variance Explained_3",col=4)
c = svd3$d^2/sum(svd3$d^2)
c[1]
```



### 2. Load the Montgomery and Pickrell eSet:
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```
### Perform the log2(data + 1) transform and subtract row means from the samples. Set the seed to 333 and use k-means to cluster the samples into two clusters. Use svd to calculate the singular vectors. 
### **Q2. What is the correlation between the first singular vector and the sample clustering indicator?**
```{r}
# log2 transform and subtract row means
edata_log2 = log2(edata + 1)
edata_centered = edata_log2 - rowMeans(edata_log2)

# use svd to calculate the singular vectors
set.seed(333)
svd1 = svd(edata_centered)

# perform k-means clustering as two clustering centers
kmeans1 = kmeans(t(edata_centered), centers=2)
cor(svd1$v[,1], kmeans1$cluster)
```



### 3. Load the Bodymap data with the following command
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
```
### Fit a linear model relating the first gene’s counts to the number of technical replicates, treating the number of replicates as a factor. Plot the data for this gene versus the covariate.
### **Q3. Can you think of why this model might not fit well?**
```{r}
# fit a linear model
lm1 = lm(edata[1,] ~ pdata_bm$num.tech.reps)

# plot the data
plot(pdata_bm$num.tech.reps,edata[1,], col=2)
abline(lm1$coeff[1], lm1$coeff[2], col=4, lwd=3)
```
### Answer for Q3:
There are very few samples with more than 2 replicates so the estimates for those values will not be very good.  



### 4. Fit a linear model relating he first gene’s counts to the age of the person and the sex of the samples. 
### **Q4. What is the value and interpretation of the coefficient for age?**
```{r}
# fit linear model
lm2 = lm(edata[1,] ~ pdata_bm$age + pdata_bm$gender)
summary(lm2)
```
### Answer for Q4:
This coefficient means that for each additional year of age, the count goes down by an average of 23.91 for a fixed sex.  



### 5. Load the Montgomery and Pickrell eSet:
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```
### Perform the log2(data + 1) transform. Then fit a regression model to each sample using population as the outcome. Do this using the lm.fit function (hint: don't forget the intercept). 
### **Q5. What is the dimension of the residual matrix, the effects matrix and the coefficients matrix?**
```{r}
# log2 transform
edata = log2(edata + 1)

# fit many regression models to each sample, using population as the outcome
mod = model.matrix(~ pdata$population)
fit = lm.fit(mod, t(edata))
names(fit)
```
```{r}
# dimension of the residual matrix
dim(fit$residuals)
```
```{r}
# dimension of the effects matrix
dim(fit$effects)
```
```{r}
# dimension of the coefficients matrix
dim(fit$coefficients)
```



### 6. Load the Montgomery and Pickrell eSet:
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```
###Perform the log2(data + 1) transform. Then fit a regression model to each sample using population as the outcome. Do this using the lm.fit function (hint: don't forget the intercept). 
### **Q6. What is the effects matrix?**
```{r}
# log2 transform
edata = log2(edata + 1)

# fit many regression models to each sample, using population as the outcome
mod = model.matrix(~ pdata$population)
fit = lm.fit(mod, t(edata))
fit$effects[1,]
```
```{r}
fit$effects[,1]
```
### Answer for Q6:
The estimated fitted values for all samples for each gene, with the values for each gene stored in the columns of the matrix. 

 	

### 7. Load the Bodymap data with the following command
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
```
### Fit many regression models to the expression data where age is the outcome variable using the lmFit function from the limma package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). 
### **Q7. What is the coefficient for age for the 1,000th gene? Make a plot of the data and fitted values for this gene. Does the model fit well?**
```{r results='hide', message=FALSE, warning=FALSE}
library(devtools) 
library(Biobase)
library(limma)
```
```{r}
# subset the expression data to the samples without missing values of age
pdata_bm1 = pdata_bm[!is.na(pdata_bm$age),]
edata_bm1 = edata[,!is.na(pdata_bm$age)]

# fit many regression models to the expression data where age is the outcome
mod = model.matrix(~ pdata_bm1$age)
fit_limma = lmFit(edata_bm1, mod)

# check the coefficient
fit_limma$coefficients[1000,]
```
```{r}
# make a plot of the 1,000th gene and fitted values
intercept = fit_limma$coefficients[1000,][1]
slope = fit_limma$coefficients[1000,][2]
x = edata_bm1[1000,]*slope+intercept
plot(x,pdata_bm1$age)
```
### Answer for Q7:
The model doesn't fit well since there are two large outlying values and the rest of the values are near zero.  



### 8. Fit many regression models to the expression data where age is the outcome variable and tissue.type is an adjustment variable using the lmFit function from the limma package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). ### **Q8. What is wrong with this model?**
```{r}
# coefficients to estimate per gene
pdata_bm$tissue.type

# data points per gene
length(x)
```
### Answer for Q8:
Since tissue.type is a factor variable with many levels, this model has more coefficients to estimate per gene (18) than data points per gene (16).



### **Q9. Why is it difficult to distinguish the study effect from the population effect in the Montgomery Pickrell dataset from ReCount?**
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
head(pdata)
```
### Answer for Q9:
The effects are difficult to distinguish because the study variable and population variable are perfectly correlated. 



### 10. Load the Bodymap data with the following command
```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
```
### Set the seed using the command set.seed(33353) then estimate a single surrogate variable using the sva function after log2(data + 1) transforming the expression data, removing rows with rowMeans less than 1, and treating age as the outcome (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). 
### **Q10. What is the correlation between the estimated surrogate for batch and age? Is the surrogate more highly correlated with race or gender?**
```{r results='hide', message=FALSE, warning=FALSE}
library(devtools)
library(Biobase)
library(sva)  
library(bladderbatch)
```
```{r echo=TRUE, message=FALSE, warning=FALSE}
# preprocess the data
pdata_bm2 = pdata_bm[!is.na(pdata_bm$age),]
edata_bm2 = edata[,!is.na(pdata_bm$age)]

# log2 transformation
edata_bm2 = log2(edata_bm2 + 1)

# remove rows with rowMeans less than 1
edata_bm2 = edata_bm2[rowMeans(edata_bm2) > 1,]

# infer batch variables with sva
mod = model.matrix(~age, data=pdata_bm2)
mod0 = model.matrix(~1, data=pdata_bm2)
sva1 = sva(edata_bm2, mod,mod0, n.sv=2)
```
```{r echo=TRUE, message=FALSE, warning=FALSE}
# correlation between surrogate for batch and age
cor(sva1$sv, pdata_bm2$age)
```
```{r}
# correlation between surrogate for batch and race
cor(sva1$sv, as.numeric(pdata_bm2$race))
```
```{r}
# correlation between surrogate for batch and gender
cor(sva1$sv, as.numeric(pdata_bm2$gender))
```
```{r}
edata_bm2 = log2(edata_bm2 + 1)

pdata_bm2 = pdata_bm[!is.na(pdata_bm$age),]
edata_bm2 = edata[,!is.na(pdata_bm$age)]

edata_bm2 = edata_bm2[rowMeans(edata_bm2) > 1,]

mod = model.matrix(~age, data=pdata_bm2)
mod0 = model.matrix(~1, data=pdata_bm2)
sva1 = sva(edata_bm2, mod,mod0, n.sv=2)

summary(lm(sva1$sv ~ pdata_bm2$age))
```



```{r}
sessionInfo()
```